\chapter[Editors Foreword]{}
\section*{Editors Foreword}

This book has a single aim: to help the reader understand the true nature of the reality of the present moment. This moment.

It consists of 9 travelogues, discussions with Acharn Sujin Boriharnwanaket, the good friend in Dhamma, a Thai Buddhist lay follower. They have taken place over the last seven years in a variety of places, mostly Thailand, South Vietnam and Sri Lanka. The various situations are used to show the relevance of the teachings to daily life. The book starts with an interview with the author about her book \textit{Abhidhamma in Daily Life}. She has also written and has published 4 other books on Abhidhamma topics.

Nina's writing is straight, uncluttered, she uses simple words to disseminate and embellish the deep meaning of Sujin's dhamma talks. She draws on her detailed knowledge of Abhidhamma texts and for precision uses some Pāli words. Quotes are used from the original Theravāda Suttas to show that the meaning of these words is consistent with the teachings of the Buddha. Her writing and Sujin's conversation help the reader understand the depth and subtlety of the teachings. They help the reader understand the realities of daily life. The reality which `appears' now.

The book is suitable for the beginner, but with a caveat. It is difficult and takes great patience over a long period to appreciate. It is an impressive piece of work but only for those who have persevered with its meaning.

I would like to thank Acharn Sujin Boriharnwanaket, Nina van Gorkom, Sarah and Jonothan Abbott and the Dhamma and Support Foundation who have helped me glimpse the depth and subtlety of this great teaching. 

With much gratitude and appreciation, I share this work with a wider audience.

Alan Weller