\part{What is Most Precious?}
\chapter{Preface}

Acharn Sujin and friends were invited again for Dhamma discussions in
Vietnam by Tam Bach and other Vietnamese friends at the end of August
and the beginning of September 2017. Before this journey Dhamma
discussions were arranged in Thailand, in Kaeng Krachan and Nakorn
Nayok. I only attended the discussions in Nakorn Nayok. Here I received
a great lesson, namely, that one never knows what will happen the next
moment. It all depends on conditions and it is beyond control. I
suddenly lost my balance and fell down. Since there was no fracture I
could get up and continue to attend the sessions. Thanks to Hang, a
Vietnamese friend who had helped me during former trips in many ways, I
could continue my journey to Vietnam in a wheelchair. Hang had special
connections so that she could arrange a wheelchair for me.

I am most grateful to her for all her kindness, consideration and
patience in constantly helping me in the wheelchair and out of it, to go
up and down on steep staircases, to overcome all the troubles and
hardship. She kept on being optimistic and seeing things from the
sunny side. Without her I could not have continued this journey and I
would have missed all the opportunities for hearing and considering the
Dhamma. Hang said all the time: ``no worry'', reminding me that we keep
on being preoccupied with ourselves when we worry about the future and
how to face difficulties. When I was annoyed about my laundry she said:
``It is past, it is all gone.'' It is very helpful to be reminded that
whatever happens are all conditioned dhammas that cannot be controlled.
This was a topic that was raised again and again during our journey. We
kept on reminding ourselves: ``All gone, all gone''.

During the discussions Acharn Sujin was assisted by Sarah and Jonothan
and they helped me and others not to be forgetful of the truth of life,
to realize what is most precious: understanding the reality of the
present moment. We are absorbed in our thinking of stories and worries
which are not real. Seeing, hearing or thinking are real, they are
realities that do not last and that are not self.
